Unsupervised training of maximum-entropy models for lexical selection in rule-based machine translation

### 2015-02-10 2015-eamt-melsa/2015-eamt-melsa.tex
down to 9
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@58800 72bbbca6-d526-0410-a7d9-f06f51895060

The remainder of the article is organised as follows: 
\todo[inline]{FMT: Add article structure}

@@

Let the probability of a translation $t$ being the translation of a
word $s$ in an SL context $c$ be $p_s(t|c)$. In principle, this value
could be estimated directly from the available corpora for every
combination of $(s, t, c)$.  This would however present two questions:
(1) how should the relevant contexts be chosen? and (2) what should be
done when the translations of words are not found in the corpus?  A
maximum-entropy model answers both of these questions. It allows the
contexts that we consider to be linguistically interesting to be
defined \emph{a priori} and then integrate these seamlessly into a
probabilistic model \citep{Manning99b}.  In answer to the second
question, a maximum-entropy model assumes nothing about what is not
present in the training data. That is, if there is no information in
the training data, then it assumes that all outcomes ---that is, all
possible translations--- are equally likely.  As previously mentioned,
the principle of maximum entropy has been applied to the problem of
lexical selection before; in particular, \cite{berger1996} cast the
problem of lexical selection in statistical MT as a classification
problem. They learn a separate maximum-entropy classifier for each SL
word form,\comment{[:we do too, integrate this later]} using SL context to
distinguish between possible translations. These classifiers are then
incorporated into the translation model of their word-based SMT
system.\comment{[how is our work different?]}  In their approach, a
classifier consists of a set of binary feature functions and
corresponding weights for each feature. 
In both \cite{berger1996} and our method, features are defined in the
form $h_k^{s}(t, c)$, where $t$
is a translation, and $c$ is an SL context. One difference is that \cite{berger1996} take $s$, $t$
and $c$ to be based on word forms, whereas in our method they are based
on lemma forms.
A feature where the Spanish word \emph{pez} (`fish' as a living animal) is seen as the
translation of \emph{arrain} (`fish') in the context \emph{arrain
  handi} `big fish' would therefore be defined as:
%\comment{FSM: I have introduced the index $k$
%  because several features are defined for a SL word $s$.} 
\begin{equation}
h_k^{\mathrm{arrain}}(t, c) = \left \{ \begin{matrix}
                 ~~1 & \mbox{if } t~ =~ \textit{pez}~ \mathrm{and}~ \textit{handi}~ \mathrm{follows}~ \textit{arrain}   \\
                 ~~0 & \mbox{otherwise} \end{matrix} \right.
\label{eq:feat-func}
\end{equation}
This feature considers a context of zero words to the left of the
problem word and one word to the right of it.

@@

Evaluation will be performed using four language pairs. These pairs
have been selected as they include languages with different
morphological complexity, and different amounts of resources available
--- although for all pairs there is a parallel corpus available, which
is used for evaluation (see Section \ref{ss:measures}). Table~\ref{table:lang-pair-stats} 
presents statistics about the development status of each of the language 
pairs chosen, including the number of bilingual dictionary entries, coverage of
test corpora and translational ambiguity of the bilingual dictionary. Translation
ambiguity is provided for both all words, and for only ambiguous open-category 
words. In the experiments, lexical-selection rules are learnt for the open categories
of adjective, noun and verb.

\begin{table*}
  \centering
  \begin{tabular}{|l|r|r|r|r|r|}
    \hline
    \multirow{2}{*}{\textbf{Pair}} & \textbf{Lexical} & \multirow{2}{*}{\textbf{Ambig.}} & \textbf{Na\"{i}ve}& \textbf{Total} & \textbf{Ambig. of}\\\
                          & \textbf{items} &                            & \textbf{coverage} & \textbf{ambig.} & \textbf{ambig. words}\\
    \hline
    \texttt{br-fr} & 27,690 & 235  &  94.47 \%  & 1.02 & 2.5 \\
    \texttt{mk-en} & 33,229 & 1,323 & 92.17 \%  & 1.37 & 2.8 \\
    \texttt{eu-es} & 19,323 & 455 & 91.70 \% & 1.08 & 2.4 \\
    \texttt{en-es} & 30,186 & 822 & 98.08 \% & 1.05 & 2.2 \\
    \hline
  \end{tabular}
  \caption{Statistics about development status of language pairs. The second
    column gives the number of entries in the bilingual dictionaries of the systems;
    the third column gives the total number of source language words with more 
    than one translation in the target language; the fourth column gives the coverage
    of the dictionaries on the test corpus (see section~\ref{sec:eval-corpus}); the fifth
    column gives the total mean ambiguity for open classes (nouns, verbs, adjectives);
    and the final column gives the mean ambiguity for ambiguous words in open classes.}
  \label{table:lang-pair-stats}
\end{table*}

@@

\begin{description}
\item[Breton--French:] Development of the Breton--French pair has been
  described in two articles \citep{tyers09b,tyers10b}. The bilingual
  dictionaries were not built with polysemy in mind from the outset,
  but some entries were added later to start work on lexical
  selection. The version used in this paper is SVN revision
  41375.\footnote{\url{https://svn.code.sf.net/p/apertium/svn/trunk/apertium-br-fr}}
%  This is the only available machine translation system between Breton
%  and French. It has been developed part-time over a number of years.

\item[Macedonian--English:] The Macedonian--English pair
  in Apertium was created specifically for the purposes of running
  lexical-selection experiments. 
% The resources reused from other pairs
%  were the English morphological dictionary from the Icelandic and
%  English pair \citep{brandt11}, the Macedonian morphological
%  dictionary and constraint grammar from the Macedonian and Bulgarian
%  pair \citep{rangelov11}, and the SETimes parallel corpus
  The lexical resources for the pair were tuned to the SETimes parallel corpus
  \citep{tyers10}.  
% The work was carried out over a period of eight
%  days, and consisted of creating the bilingual dictionary and
%  transfer rules. The bilingual dictionary was created by tagging both
%  sides of the parallel corpus, word-aligning them with {\sc giza++}
%  \citep{och03a} and extracting the probabilistic lexicon. Entries
%  from this lexicon were checked manually according to frequency and
%  included in the bilingual dictionary of the machine translation
%  system. 
  The most probable entry from automatic word alignment of this 
  corpus using {\sc giza++} \citep{och03a} was checked to ensure that it was 
  an adequate translation, and if so marked as the 
  default.\footnote{A bilingual dictionary in Apertium
    \citep{forcada2011apertium} may contain more than one possible
    target-language translation for a given source-language
    word. Where there is more than one possible translation, the
    person who is writing the dictionary chooses the most general or
    most frequent translation among the set of possible translations
    and marks it as the \emph{linguistic default} translation.} As a
  result of attempting to include all possible translations, the
  average number of translations per word is much higher (see table~\ref{table:lang-pair-stats})
than in other pairs.  
% The transfer rules were written
%  by hand to treat translation problems in the corpus. The version of
%  the software used in this paper is SVN revision
%  41476.\footnote{\url{https://svn.code.sf.net/p/apertium/svn/trunk/apertium-mk-en}}.
\item[Basque--Spanish:] The development of the Basque to Spanish pair
  is described in \cite{ginesti09}. 
% The bilingual dictionary was taken
%  from the free/open-source Matxin system \citep{mayor2011}. 
  Alternative translations were included in the bilingual
  dictionary.% , but non-default translations were marked with a
%  direction restriction\comment{FSM: This needs to be carefully
%    explained to avoid confusing the reader.}.\footnote{The bilingual
%    dictionaries in Apertium are bidirectional. They contain entries
%    which can be used for translating both from language $A$ to
%    language $B$ and vice versa. A direction restriction is used to
%    limit the entry to being used for translation in one translation
%    direction.} For lexical-selection experiments these direction
%  restrictions were removed.

@@

\subsubsection{Machine translation performance} 
\comment{This part of the text needs compressing and merging} This is
an extrinsic evaluation, which ideally would test how much the system
improves as regards an approximate measurement of final translation
quality in a real system. For this task, we use the widely-used {\sc
  bleu} metric (see Section~\ref{sec:eval-metrics}). This is not ideal
for evaluating the task of a lexical selection module as the
performance of the module will depend greatly on (a) the coverage of
the bilingual dictionaries of the RBMT system in question, and (b) the
number of reference translations. It is also worth noting that
successful lexical selections may not lead to successful translations
due to inadequate transfer of morphological features. For example, if
we translate the Spanish phrase \emph{La gente le dice que no venga}
`The people tell him not to come', where \emph{decir} can be
translated as `say' or `tell' in English, and our machine translation
system generates `The people \emph{tells} him not to come', it would
be counted the same as if it were `The people \emph{says} him not to
come', although the lexical selection made was more
adequate. Generation errors --- where a word is found in the bilingual
dictionary, but not in the target language morphological dictionary
--- may also lead to the same problem.  The {\sc bleu} metric \citep{papineni02} is
included only as it is commonly used to evaluate MT systems.

@@

The ideal situation would be to have, as described by
\cite{zinovjeva2000}, a hand-annotated evaluation corpus for testing
the performance of the lexical-selection module. That is, the output
of the lexical transfer module which had been disambiguated by hand by
one or more human annotators. As this did not exist for any language
pair, we decided to automatically annotate a test set for each
language pair using a process similar to that described by
\cite{vickrey05}. The decision was made to automatically annotate
instead of hand annotate for a number of reasons. Firstly, the time
involved in hand-annotating approximately 30 pages of text, per
language pair, would be prohibitive. Secondly, linguists were not
available for all of the language pairs. One disadvantage
to this method is that the set of
translations for each ambiguous word may not be complete as it will
only include the translation which is found in the corresponding
sentence in the parallel corpus. Further disadvantages are that 
not all sentence pairs in the parallel corpus will have translations
that are able to be generated by the MT system (for
example the reference translation has a word which is not found in
the bilingual dictionaries on the MT system).

@@

\begin{itemize}
\item \textbf{Linguist-chosen defaults}. 
%! \comment{FSM: This should go to the section explaining Apertium.}  
  A bilingual dictionary in an
  Apertium language pair contains correspondences between lexical
  forms. The dictionaries allow many lexical forms to translate to one
  lexical form. For example, \emph{ahizpa} and \emph{arreba} in Basque
  both translate to \emph{hermana} `sister' in Spanish.\footnote{The
    difference being that \emph{ahizpa} is a sister of a female, and
    \emph{arreba} is a sister of a male.} But a single lexical form
  may not have more than one translation without further
  processing. If there are many possible translations of a lexical
  form, then one must be marked as the \emph{default} translation. For
  example, \emph{pisu} in Basque translates to \emph{peso} `weight'
  and \emph{piso} `flat' in Spanish. If the translation correspondence
  \emph{pisu $\rightarrow$ peso} is perceived as the most frequent or
  the most general, it is marked as the default by the linguist who is
  creating the bilingual dictionary.

\item \textbf{Oracle}.  The results for the oracle system are those
  achieved by passing the automatically annotated reference
  translation through the rest of the modules of the MT system. This
  is included to give an idea of the upper bound for the performance
  of the lexical-selection module. If all of the lexical choices were
  made in accordance with the reference, this is the result that would
  be achieved.

\item \textbf{Target language model} (TLM). One method of lexical
  selection is to use the existing machine translation system to
  generate all the possible translations for an input sentence, and
  then score these translations \emph{on-line} on a model of the
  target language. The highest scoring sentence is then picked as
  output. This is the method used by %@@@ \cite{carbonell06} and
  \cite{melero07a}.  We used a five-gram language model of surface
  forms generated from the target-language side of the parallel
  corpus. %\comment{FSM: Why not a much larger corpus? (I reviewer may ask)}
%(see section~\ref{sec:eval-corpus}). 
  This was the best performing system when we compared different
  approaches, but is impractical for real-world MT because the number
  of translations to perform grows exponentially with the length of
  the sentence.  It represents the best that can be achieved using the
  current systems without access to a parallel corpus. In our work, it
  was implemented with the IRST language modelling toolkit
  \cite{federico08a}.
\end{itemize}

>>>>

%The remainder of the article is organised as follows: 
%\todo[inline]{FMT: Add article structure}

@@

Let the probability of a translation $t$ being the translation of a
word $s$ in an SL context $c$ be $p_s(t|c)$. In principle, this value
could be estimated directly from the available corpora for every
combination of $(s, t, c)$.  This would however present two questions:
(1) how should the relevant contexts be chosen? and (2) what should be
done when the translations of words are not found in the corpus?  A
maximum-entropy model answers both of these questions. It allows the
contexts that we consider to be linguistically interesting to be
defined \emph{a priori} and then integrate these seamlessly into a
probabilistic model \citep{Manning99b}.  In answer to the second
question, a maximum-entropy model assumes nothing about what is not
present in the training data. That is, if there is no information in
the training data, then it assumes that all outcomes ---that is, all
possible translations--- are equally likely.  As previously mentioned,
the principle of maximum entropy has been applied to the problem of
lexical selection before; in particular, \cite{berger1996} cast the
problem of lexical selection in statistical MT as a classification
problem. They learn a separate maximum-entropy classifier for each SL
word form,
%\comment{[:we do too, integrate this later]} 
using SL context to
distinguish between possible translations. These classifiers are then
incorporated into the translation model of their word-based SMT
system.
%\comment{[how is our work different?]}  
In their approach, a
classifier consists of a set of binary feature functions and
corresponding weights for each feature. 
In both \cite{berger1996} and our method, features are defined in the
form $h_k^{s}(t, c)$, where $t$
is a translation, and $c$ is an SL context. One difference is that \cite{berger1996} take $s$, $t$
and $c$ to be based on word forms, whereas in our method they are based
on lemma forms.
A feature where the Spanish word \emph{pez} (`fish' as a living animal) is seen as the
translation of \emph{arrain} (`fish') in the context \emph{arrain
  handi} `big fish' would therefore be defined as:
%\comment{FSM: I have introduced the index $k$
%  because several features are defined for a SL word $s$.} 
\begin{equation}
h_k^{\mathrm{arrain}}(t, c) = \left \{ \begin{matrix}
                 ~~1 & \mbox{if } t~ =~ \textit{pez}~ \mathrm{and}~ \textit{handi}~ \mathrm{follows}~ \textit{arrain}   \\
                 ~~0 & \mbox{otherwise} \end{matrix} \right.
\label{eq:feat-func}
\end{equation}
This feature considers a context of zero words to the left of the
problem word and one word to the right of it.

@@

Evaluation will be performed using four language pairs. These pairs
have been selected as they include languages with different
morphological complexity, and different amounts of resources available
--- although for all pairs there is a parallel corpus available, which
is used for evaluation (see Section \ref{ss:measures}). 
% Table~\ref{table:lang-pair-stats} 
%presents statistics about the development status of each of the language 
%pairs chosen, including the number of bilingual dictionary entries, coverage of
%test corpora and translational ambiguity of the bilingual dictionary. Translation
%ambiguity is provided for both all words, and for only ambiguous open-category 
%words. In the experiments, lexical-selection rules are learnt for the open categories
%of adjective, noun and verb.

% \begin{table*}
%   \centering
%   \begin{tabular}{|l|r|r|r|r|r|}
%     \hline
%     \multirow{2}{*}{\textbf{Pair}} & \textbf{Lexical} & \multirow{2}{*}{\textbf{Ambig.}} & \textbf{Na\"{i}ve}& \textbf{Total} & \textbf{Ambig. of}\\\
%                           & \textbf{items} &                            & \textbf{coverage} & \textbf{ambig.} & \textbf{ambig. words}\\
%     \hline
%     \texttt{br-fr} & 27,690 & 235  &  94.47 \%  & 1.02 & 2.5 \\
%     \texttt{mk-en} & 33,229 & 1,323 & 92.17 \%  & 1.37 & 2.8 \\
%     \texttt{eu-es} & 19,323 & 455 & 91.70 \% & 1.08 & 2.4 \\
%     \texttt{en-es} & 30,186 & 822 & 98.08 \% & 1.05 & 2.2 \\
%     \hline
%   \end{tabular}
%   \caption{Statistics about development status of language pairs. The second
%     column gives the number of entries in the bilingual dictionaries of the systems;
%     the third column gives the total number of source language words with more 
%     than one translation in the target language; the fourth column gives the coverage
%     of the dictionaries on the test corpus (see section~\ref{sec:eval-corpus}); the fifth
%     column gives the total mean ambiguity for open classes (nouns, verbs, adjectives);
%     and the final column gives the mean ambiguity for ambiguous words in open classes.}
%   \label{table:lang-pair-stats}
% \end{table*}

@@

\begin{description}
\item[Breton--French:] Development of the Breton--French pair has been
  described in \citep{tyers10b}. The bilingual % tyers09b
  dictionaries were not built with polysemy in mind from the outset,
  but some entries were added later to start work on lexical
  selection. The version used in this paper is SVN revision
  41375.\footnote{\url{https://svn.code.sf.net/p/apertium/svn/trunk/apertium-br-fr}}
%  This is the only available machine translation system between Breton
%  and French. It has been developed part-time over a number of years.

\item[Macedonian--English:] The Macedonian--English pair
  in Apertium was created specifically for the purposes of running
  lexical-selection experiments. 
% The resources reused from other pairs
%  were the English morphological dictionary from the Icelandic and
%  English pair \citep{brandt11}, the Macedonian morphological
%  dictionary and constraint grammar from the Macedonian and Bulgarian
%  pair \citep{rangelov11}, and the SETimes parallel corpus
  The lexical resources for the pair were tuned to the SETimes parallel corpus
  \citep{tyers10}.  
% The work was carried out over a period of eight
%  days, and consisted of creating the bilingual dictionary and
%  transfer rules. The bilingual dictionary was created by tagging both
%  sides of the parallel corpus, word-aligning them with {\sc giza++}
%  \citep{och03a} and extracting the probabilistic lexicon. Entries
%  from this lexicon were checked manually according to frequency and
%  included in the bilingual dictionary of the machine translation
%  system. 
  The most probable entry from automatic word alignment of this 
  corpus using {\sc giza++} \citep{och03a} was checked to ensure that it was 
  an adequate translation, and if so marked as the 
  default.\footnote{A bilingual dictionary in Apertium
    \citep{forcada2011apertium} may contain more than one possible
    target-language translation for a given source-language
    word. Where there is more than one possible translation, the
    person who is writing the dictionary chooses the most general or
    most frequent translation among the set of possible translations
    and marks it as the \emph{linguistic default} translation.} As a
  result of attempting to include all possible translations, the
  average number of translations per word is much higher % (see table~\ref{table:lang-pair-stats})
than in other pairs.  
% The transfer rules were written
%  by hand to treat translation problems in the corpus. The version of
%  the software used in this paper is SVN revision
%  41476.\footnote{\url{https://svn.code.sf.net/p/apertium/svn/trunk/apertium-mk-en}}.
\item[Basque--Spanish:] The development of the Basque to Spanish pair
  is described in \cite{ginesti09}. 
% The bilingual dictionary was taken
%  from the free/open-source Matxin system \citep{mayor2011}. 
  Alternative translations were included in the bilingual
  dictionary.% , but non-default translations were marked with a
%  direction restriction\comment{FSM: This needs to be carefully
%    explained to avoid confusing the reader.}.\footnote{The bilingual
%    dictionaries in Apertium are bidirectional. They contain entries
%    which can be used for translating both from language $A$ to
%    language $B$ and vice versa. A direction restriction is used to
%    limit the entry to being used for translation in one translation
%    direction.} For lexical-selection experiments these direction
%  restrictions were removed.

@@

\subsubsection{Machine translation performance} 
%! \comment{This part of the text needs compressing and merging} 
This is
an extrinsic evaluation, which ideally would test how much the system
improves as regards an approximate measurement of final translation
quality in a real system. For this task, we use the widely-used {\sc
  bleu} metric (see Section~\ref{sec:eval-metrics}). This is not ideal
for evaluating the task of a lexical selection module as the
performance of the module will depend greatly on (a) the coverage of
the bilingual dictionaries of the RBMT system in question, and (b) the
number of reference translations. It is also worth noting that
successful lexical selections may not lead to successful translations
due to inadequate transfer of morphological features. 
% For example, if
% we translate the Spanish phrase \emph{La gente le dice que no venga}
% `The people tell him not to come', where \emph{decir} can be
% translated as `say' or `tell' in English, and our machine translation
% system generates `The people \emph{tells} him not to come', it would
% be counted the same as if it were `The people \emph{says} him not to
% come', although the lexical selection made was more
% adequate. Generation errors --- where a word is found in the bilingual
% dictionary, but not in the target language morphological dictionary
% --- may also lead to the same problem.  
The {\sc bleu} metric \citep{papineni02} is
included only as it is commonly used to evaluate MT systems.

@@

The ideal situation would be to have, as described by
\cite{zinovjeva2000}, a hand-annotated evaluation corpus for testing
the performance of the lexical-selection module. That is, the output
of the lexical transfer module which had been disambiguated by hand by
one or more human annotators. As this did not exist for any language
pair, we decided to automatically annotate a test set for each
language pair using a process similar to that described by
\cite{vickrey05}.

% The decision was made to automatically annotate
% instead of hand annotate for a number of reasons. Firstly, the time
% involved in hand-annotating approximately 30 pages of text, per
% language pair, would be prohibitive. Secondly, linguists were not
% available for all of the language pairs. One disadvantage
% to this method is that the set of
% translations for each ambiguous word may not be complete as it will
% only include the translation which is found in the corresponding
% sentence in the parallel corpus. Further disadvantages are that 
% not all sentence pairs in the parallel corpus will have translations
% that are able to be generated by the MT system (for
% example the reference translation has a word which is not found in
% the bilingual dictionaries on the MT system).

@@

\begin{itemize}
\item \textbf{Linguist-chosen defaults}. 
%! \comment{FSM: This should go to the section explaining Apertium.}  
  A bilingual dictionary in an
  Apertium language pair contains correspondences between lexical
  forms. The dictionaries allow many lexical forms to translate to one
  lexical form. For example, \emph{ahizpa} and \emph{arreba} in Basque
  both translate to \emph{hermana} `sister' in Spanish.\footnote{The
    difference being that \emph{ahizpa} is a sister of a female, and
    \emph{arreba} is a sister of a male.} But a single lexical form
  may not have more than one translation without further
  processing. If there are many possible translations of a lexical
  form, then one must be marked as the \emph{default} translation. For
  example, \emph{pisu} in Basque translates to \emph{peso} `weight'
  and \emph{piso} `flat' in Spanish. If the translation correspondence
  \emph{pisu $\rightarrow$ peso} is perceived as the most frequent or
  the most general, it is marked as the default by the author of the dictionary. % linguist who is
%  creating the bilingual dictionary.

\item \textbf{Oracle}.  The results for the oracle system are those
  achieved by passing the automatically annotated reference
  translation through the rest of the modules of the MT system. This
  is included to give an idea of the upper bound for the performance
  of the lexical-selection module. % If all of the lexical choices were
%  made in accordance with the reference, this is the result that would
%  be achieved.

\item \textbf{Target language model} (TLM). One method of lexical
  selection is to use the existing machine translation system to
  generate all the possible translations for an input sentence, and
  then score these translations \emph{on-line} on a model of the
  target language. The highest scoring sentence is then picked as
  output. This is the method used by %@@@ \cite{carbonell06} and
  \cite{melero07a}.  
% We used a five-gram language model of surface
%   forms generated from the target-language side of the parallel
%   corpus. %\comment{FSM: Why not a much larger corpus? (I reviewer may ask)}
% %(see section~\ref{sec:eval-corpus}). 
%   This was the best performing system when we compared different
%   approaches, but is impractical for real-world MT because the number
%   of translations to perform grows exponentially with the length of
%   the sentence.  It represents the best that can be achieved using the
%   current systems without access to a parallel corpus. In our work, it
%   was implemented with the IRST language modelling toolkit
%   \cite{federico08a}.
\end{itemize}
